Repeatability of Pre-Flashover Fire Patterns on Gypsum Wallboard

### 2016-10-02 Projects/Madrzykowski_Thesis/Report/Madrzykowski_Thesis.tex
Syncing versions with added Fig 2.7

Plots of the total heat flux for each of the fuels are presented in Fig.~\ref{Flux}. The repeatability of the total heat flux values varied in a manner similar to the heat release rates, in that the natural gas fires provided a steady state value while the gasoline and the polyurethane foam fires generated a fire curve that has periods of growth and decay.  The average peak total heat flux for the natural gas fires were approximately 2.0~kW/m$^2$.  The average peak values for the gasoline and polyurethane foam were approximately 3.5~kW/m$^2$ and 2.5~kW/m$^2$ respectively.  The values are only presented for the experiments which had two heat flux gauges installed as described in the instrumentation section above.

\begin{figure}[p]
  \centering
  \includegraphics[width=.635\columnwidth]{../Figures/HF_NG}\\
  \includegraphics[width=.635\columnwidth]{../Figures/HF_GAS}\\
  \includegraphics[width=.635\columnwidth]{../Figures/HF_PUF}\\
  \caption[Total heat flux versus time for natural gas, gasoline, and foam fires]{Total heat flux versus time for natural gas, gasoline, and foam fires.}
  \label{Flux}
\end{figure}

@@

The cone calorimeter experiments generated heat release are per unit area values ranging from a low of approximately 160 kW/m$^2$ at an incident heat flux of 35 kW/m$^2$ to approximately 230 kW/m$^2$ at an incident heat flux of 75 kW/m$^2$.  The average total heat released during each burn was 2.8 MJ/m$^2$.

\begin{table}
	\centering
	\begin{tabular}{|c|c|}
		\hline Incident Heat Flux & Heat Release Rate per Unit Area    \\
		\hline (kW/m$^2$) & (kW/m$^2$)  \\ \hline
		\hline 35 	& 163 	\\
		\hline 50	& 179   \\
		\hline 70	& 255  	\\
		\hline 75	& 231 	\\
		\hline
	\end{tabular}
	\caption[Painted gypsum wallboard time heat release rate per unit area results]{Painted gypsum wallboard heat release rate per unit area results from the cone calorimeter. Each value is based on an average of three tests.}
	\label{tab:Gypsum wallboard_HRRA}
\end{table}

Seven replicate experiments were conducted in the radiant flooring panel apparatus.  The mean burn length was 432~mm with a 95$\%$ confidence limit of $\pm$18~mm. Comparing this burn length range, 414~mm to 450~mm, to the incident heat flux curve provided a critical heat flux of approximately 8.9~kW/m$^2$ with a 95$\%$ confidence limit of $\pm$~0.6~kW/m$^2$.  Photographs showing a radiant flooring panel test underway and sample nearing the flameout position are provided in Figure~\ref{FRP_Test_Photos}.

While the vertical radiant panel did not provide ignition data or heat release rate values it did provide information on the fire effect development on the painted gypsum wallboard samples with respect to heat flux.  Referring again to the five fire effect categories listed by DeHann~\cite{DeHaan:2012} and examining the Figure~\ref{RPGB_SET} we see surface thermal effects of discoloration on the painted paper surface of the gypsum wallboard samples at both the 10 and 12.5~kW/m$^2$ incident heat flux exposures.  The samples exposed to 15~kW/m$^2$ exhibit charring and given the thin nature of the painted paper surface penetration was evidenced on two of the three sample.  The sample of the right also shows signs of the painted paper being consumed.  The samples exposed to 17.5~kW/m$^2$ and above all show signs of significant levels of painted paper consumption.  As the heat flux was increased, the area of damage also increased. Even though the incident heat flux values were lower than the characteristic value measured at the center of the sample, as the heat flux in the center increased heat flux at the edges also increased above 15~kW/m$^2$. Hence the damaged area increased outward toward the edges. It would appear in this test series, that 15~kW/m$^2$ is right on the threshold of painted paper penetration and consumption.

DeHaan~\cite{DeHaan:2012} categorizes the fire effects that form patterns as follows:
\begin{enumerate}
	\item Surface deposits -– no irreversible effect on surface
	\item Surface thermal effects –- physical change such a discoloration or melting
	\item Charring -– evidence of surface burning
	\item Penetration -– charring below the surface
	\item Consumption –- loss of surface material, charring throughout
\end{enumerate}

Each apparatus provided similar but different thermal exposure conditions and each test method has a specific parameter or parameters that they were designed to measure for. Currently there is no specific protocol for a fire model user to follow to develop input data.  Nor is there a ``standard'' catalog of fire model inputs for a given material.  As a result the literature is sourced for information and users may need VII and VIII or to evaluate the validity of the calculations in those chapters.

>>>>

Plots of the total heat flux for each of the fuels are presented in Fig.~\ref{Flux}. The repeatability of the total heat flux values varied in a manner similar to the heat release rates, in that the natural gas fires provided a steady state value while the gasoline and the polyurethane foam fires generated a fire curve that has periods of growth and decay.  The average peak total heat flux for the natural gas fires were approximately 2.0~kW/m$^2$.  The average peak values for the gasoline and polyurethane foam were approximately 3.5~kW/m$^2$ and 2.5~kW/m$^2$ respectively.  The values in Fig.~\ref{Flux} are only presented for the experiments which had two heat flux gauges installed at 0.95~m from the centerline of the burners and 0.5~m above the top edge of the burners.

In the initial set of heat release rate experiments another heat flux gauge was used to measure the total heat flux at a location 2~m from the vertical centerline of the burners and 0.5~m above the top of the burners.  At this location the average peak total heat flux measured for the natural gas fires was approximately 0.42~kW/m$^2$.  The average peak values for the gasoline and polyurethane foam were approximately 0.74~kW/m$^2$ and 0.30~kW/m$^2$ respectively. Graphs of the heat flux time histories from the 2~m position are given in Figure~\ref{Flux_2m}

\begin{figure}[p]
  \centering
  \includegraphics[width=.635\columnwidth]{../Figures/HF_NG}\\
  \includegraphics[width=.635\columnwidth]{../Figures/HF_GAS}\\
  \includegraphics[width=.635\columnwidth]{../Figures/HF_PUF}\\
  \caption[Total heat flux versus time for natural gas, gasoline, and foam fires at 0.95~m]{Total heat flux versus time for natural gas, gasoline, and foam fires at a position 0.95~m from the vertical centerline of the burner.}
  \label{Flux}
\end{figure}

\begin{figure}[p]
  \centering
  \includegraphics[width=.635\columnwidth]{../Figures/HF_NG_2m}\\
  \includegraphics[width=.635\columnwidth]{../Figures/HF_Gas_2m}\\
  \includegraphics[width=.635\columnwidth]{../Figures/HF_PUF_2m}\\
  \caption[Total heat flux versus time for natural gas, gasoline, and foam fires at 2~m]{Total heat flux versus time for natural gas, gasoline, and foam fires at a position 2~m from the vertical centerline of the burner.}
  \label{Flux_2m}
\end{figure}

@@

While the vertical radiant panel did not provide ignition data or heat release rate values it did provide information on the fire effect development on the painted gypsum wallboard samples with respect to heat flux.  Referring again to the five fire effect categories listed by DeHann~\cite{DeHaan:2012} and examining the Figure~\ref{RPGB_SET} we see surface thermal effects of discoloration on the painted paper cover of the gypsum wallboard samples at both the 10 and 12.5~kW/m$^2$ incident heat flux exposures.  The samples exposed to 15~kW/m$^2$ exhibit charring and given the thin nature of the painted paper surface penetration was evidenced on two of the three sample.  The 15~kW/m$^2$ sample on the right also shows signs of the painted paper being consumed.  The samples exposed to 17.5~kW/m$^2$ and above all show signs of significant levels of painted paper consumption.  As the heat flux was increased, the area of damage also increased. Even though the incident heat flux values were lower than the characteristic value measured at the center of the sample, as the heat flux in the center increased heat flux at the edges also increased above 15~kW/m$^2$. Hence the damaged area increased outward toward the edges. It would appear in this test series, that 15~kW/m$^2$ heat flux exposure is on the threshold of painted paper penetration and consumption stage.

Each apparatus provided a thermal exposure to painted gypsum wallboard samples and an array of data was generated including: time to ignition, ignition temperature, heat release rate per unit area, and critical heat flux.  Each test method has a specific parameter or parameters that they were designed to measure for. As a result, the data presented in this chapter provides a wide range of values that can not be taken as a single number an applied.  In other words the data must be taken in context with how it was generated and analyzed accordingly. This is a challenge for fire investigators, analysts, and engineers alike.

The nature of the bench scale experimental data does not allow one to simply pick a value from a table an directly applying it as an input to a fire model.   Currently there is no specific protocol for a fire model user to follow to develop input data.  For example, there is no guidance on how to conduct a series of experiments to generate all of the data needed for the models.  Using the gypsum wall board as an example, material propeties such as density, specific heat, and thermal conductivity change over time as a result of the thermal exposure~\cite{Schroeder:1999,McGraw:1999}.  Of course this behavior is not unique to gypsum wallboard. The ideal case for the fire investigation user community would be a ``standard'' catalog of fire model inputs for a given material under a wide range of conditions.

There are other test methods that were not utilized here.  The lateral ignition and flame spread apparatus can be used to generate time to ignition and ignition temperature data in addition to critical heat flux for flame spread data.  The LIFT also considers time into the data where as the radiant flooring panel does not.  Further the LIFT and the cone calorimeter apparatus can be conducted in both horizontal and vertical orientation to better represent the end use of the sample material.  Thermogravimetric analysis (TGA) is another type of test which can be used to measure the mass loss rate of a material as a function of time and temperature.  Differential scanning calorimetry (DSC) is a test method which can be used to measure the specific heat capacity and in way phase change.  The TGA and DSC use samples that are on the order of a few grams.  This can make testing with a composite material as well as interpreting the results of a composite material difficult.

This chapter was anticipated to provide data that that correlated better than it does, although this should not have been a surprise given the wide range of thermal exposures, 10 to 75~kW/m$^2$.  The ignition temperatures measured in the cone calorimeter experiments conducted as part of this side do not align with the data in the published literature but the reason for the 200~$^{\circ}$C could not be resolved other than issues with thermocouple mounting, the use of paint, or changes to the gypsum board material properties over longer exposure times.
